Future challenges for Research Software Engineering

### 2025-05-12 future_challenges_for_RSEng.tex
syncronize structure

Research Software Engineering is currently an emerging field,
with an increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in imagining the future.
In the following sections, we consider the evolution of the RSE definition,
the development of RSE education,
the impact of automation and generative artificial intelligence in RSE work,
the emerging ethical considerations,
and the transformations expected in the research organizations.
We then attempt to make long-term predictions
and we conclude the paper with our wishes and recommendations.

@@

It is already difficult to structure the spectrum of RSE tasks and responsibilities today (FIXME: Willi Hasselbrings classification paper?)
and it is even more hopeless to structure the aspects that will become relavant for the RSE community in the future.
Nevertheless, we decided to consider
\begin{itemize}
\item The evolution of the term RSE,
\item How education is reshaped by the digital transformation,
\item What the impact of AI to the work of RSEs will look like,
\item Changes to the Academic Institutions,
\item Impact of alternate career paths for RSEs and other RTPs (\emph{Research Technical Professional}s).
\item Development of the RSE community,
\item Ethics and social considerations,
\end{itemize}

>>>>

Research Software Engineering is currently an emerging field,
with an increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in imagining the future.
In the following sections, we consider the evolution of the RSE definition,
the development of RSE education,
the impact of automation and generative artificial intelligence in RSE work,
the emerging ethical considerations,
and the transformations expected in the research organizations,
and if there will be other in academia covered by new RTPs (\emph{Research Technical Professional}s).
We conclude the paper with a summary and with our wishes and recommendations.

@@

It is already difficult to structure the spectrum of RSE tasks and responsibilities today (FIXME: Willi Hasselbrings classification paper?)
and it is even more hopeless to structure the aspects that will become relavant for the RSE community in the future.
Nevertheless, we decided to consider
\begin{itemize}
\item The evolution of the term RSE,
\item Development of the RSE community,
\item What the impact of AI to the work of RSEs will look like,
\item the impact of Complexity and Usability in Research Software
\item Ethics and social considerations,
\item How education is reshaped by the digital transformation,
\item Changes to the Academic Institutions,
\item Impact of alternate career paths for RSEs and other RTPs.
\end{itemize}
